Co-methylation review

### 2015-04-15 Rmarkdown/comethylation_review.Rmd
Figure fixing.
A few TODOs
Working through edits to co-methylation review

\section{Introduction}

Co-methylation is the dependence structure of DNA methylation data. This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation. In this chapter I critically review these definitions and put them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}. I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}. I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.

@@

I define co-methylation as the dependence structure of DNA methylation data. This broad definition encompasses several concepts that have variously been described as 'co-methylation', or similar. My definition is broader than "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha} or "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; it also includes the "correlation of two [loci] across many sample" \citep{Akulenko:2013fl} and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}. A first important point to emphasise is that there exists both within-sample co-methylation and between-sample co-methylation.

Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample. Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated \citep[e.g.,][]{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010dr, Landan:2012kp}. Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context.

@@

Between-sample co-methylation is the relationship between methylation levels at a pair of loci across a set of samples[^concordance]. For example, \citet{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples. Also note, however, that measures of within-sample co-methylation can themselves be compared between samples, which is discussed at the end of this chapter.

[^concordance]: One measure of between-sample correlation of methylation levels that does not fall under my definition of co-methylation is the following: Consider a pair of samples, $j$ and $j'$, and compute the correlation between the set of $\beta$-values for sample $j$ against sample $j'$, i.e., $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$. This measure is often reported as evidence for the "concordance" of methylation levels between replicates, but is not what I consider a form of co-methylation.

There have been several attempts to analyse co-methylation, most with the aim of better understanding the biological processes that lay down and regulate DNA methylation. An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation. For example, we might identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci. This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays.

@@

\section{Correlations of aggregate measurements of methylation}
\label{sec:previous_cor_of_agg_methylation}

To date, most analyses of co-methylation have been based on analysing correlations of aggregate levels of methylation, such as $\beta$-values, within individual samples \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}. These analyses have an implicit assumption of stationarity, that is, that the correlation between two loci depends only on the distance between them and not the actual position of the loci in the genome.

@@

\citet{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples. Of particular relevance to a discussion of co-methylation, \citet{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \citep{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \citep{Hansen:2011gu}. \citet{Eckhardt:2006gh} performed PCR-amplicon bisulfite-sequencing of $2,524$ amplicons on human chromosomes $6$, $20$ and $22$. PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays. Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.

As part of their analysis, the authors analyse co-methylation, which they define as "the relationship between the degree of methylation over distance". They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $>2,000$ bp" and reference the plot reprinted in Figure \ref{fig:Eckhardt} in support of this claim. In Figure \ref{fig:Eckhardt}, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp, which is not a correlation in the usual sense of the word. Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed[^beck].

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{../figures/ng1909-F3a.jpg}
\caption{Correlation of DNA methylation with spatial distance. (a) Correlation between comethylation and spatial distance. Orange dots represent CpG methylation values aggregated and averaged over $25,000$ individual measurements. Gray dots represent CpG methylation values based on resampling of random CpG positions. Blue dots indicate CpG methylation values based on resampling of amplicon positions. At distances $>1,000$ bp, we did not detect any correlation between CpG methylation and spatial distance. Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Eckhardt:2006gh}), copyright (2006)}
\label{fig:Eckhardt}
\end{figure}

[^beck]: An email discussion with the senior author, Stephan Beck, was unable to clarify the definition or method.

@@

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{i, h}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from distinct samples with different genome-wide average methylation levels. This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{\ast}$ as follows:

@@

This result is proven in Appendix \ref{app:bernoulli_equal}.

@@

1. $PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.

@@

Quite apart from what exactly is plotted in Figure \ref{fig:Eckhardt}, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis. Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure \ref{fig:Eckhardt} does not represent any one sample but a pool of samples.

@@

\citet{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing. They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis Thaliana_. They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis]. The autocorrelation computes, for a set of distances ($0-5,000$ nt), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance. To emphasise, I believe this analysis uses all pairs of methylation loci, that is, regardless of how many intervening methylation loci there are for each pair.

@@

The autocorrelation plots showed two periodicities, one of $\approx 167$ nucleotides and the other of $10$ nucleotides (CHH only). Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier analysis of the autocorrelation signal.

[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}

The $167$ nucleotide periodicity is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al. hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.

The $10$ nucleotide periodicity is equal to the length of one helical DNA turn. From this result Cokus et al. hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites $8-10$ nucleotides apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_). The $10$ nucleotide CHH-only periodicity in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})

[^comethylation_def2]: This is also referred to as "co-methylation" by \citet{Jurkowska:2011dt}, here meaning the methylation of multiple loci in a single event. \citet{Jurkowska:2011dt} reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.

\subsection{\citet{Chodavarapu:2010iqa}\label{sec:chodavarapu}

\citet{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (_HSF1_). The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq. The nucleosome start sites anchor the methylation observations to a common grid and this plot is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes. A $10$ nucleotide periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}. Again, the observed periodicity is confirmed by analysing a Fourier transform of this correlation signal.

In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the _HSF1_ sample, which confirmed the $10$ nucleotide periodicity. In fact, these results revealed that the $10$ nucleotide is common to all methylation types, including symmetric CpG methylation, at least for nucleosome-bound DNA. This led \citet{Chodavarapu:2010iqa} to discard their earlier hypothesis as to the source of the $10$ nucleotide periodicity \citep{Cokus:2008fc} in favour of one that "nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.

@@

\citet{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation". The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", citing their results as being consistent with those in \citet{Eckhardt:2006gh}. In addition to this genome-wide analysis, \citet{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements. Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.

@@

Figure \ref{fig:Li6} plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp. Based on the "$n = 18,936,995$" quoted in the header of panel (a) of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "distance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs. If all pairs were used this $n$ would be much larger since there are more than twenty million CpGs on one strand of the haploid human genome. To emphasise, the correlation shown by \citet{Li:2010fb} is not the same as the autocorrelation computed in \citet{Cokus:2008fc} since \citet{Li:2010fb} only use pairs of adjacent CpGs whereas \citet{Cokus:2008fc} use all pairs of CpGs, regardless of the number of intervening CpGs.

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{../figures/{journal.pbio.1000533.s006}.pdf}
\caption{Co-methylation patterns for different genomic features. A reproduction of Supplementary Figure 6 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
\label{fig:Li6}
\end{figure}

The genome-wide analysis correlation plot contains a periodicity of approximately $170$ nucleotides, very similar to the $167$ nucleotide period identified in _Arabidopsis_ by \citet{Cokus:2008fc} (reproduced in Figure \ref{fig:Li7}). This is further evidence that nucleosome spacing effects the strength of co-methylation. However, \citet{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ nucleotide period reported by \citet{Cokus:2008fc} and \citet{Chodavarapu:2010iqa}.

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{../figures/Figure_S7.pdf}
\caption{Transformation of methylation correlation at nearby CpG cytosines. A reproduction of Supplementary Figure 7 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
\label{fig:Li7}
\end{figure}

@@

Figure \ref{fig:Lyko} reproduces the results for the co-methylation analysis. Note that the maximum value of the "CpG autocorrelation" is less than $0.015$ (panel (A) of Figure \ref{fig:Lyko}). I strongly suspect that the autocorrelation has been incorrectly computed. Specifically, what I believe has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been artificially assigned a $\beta$-value of zero. It does not make sense to include non-CpG positions, particularly non-cytosines that can never be methylated, in an analysis of CpG co-methylation.

>>>>

__TODO: Review co-methylation in "Bell, J. T. et al. DNA methylation patterns associate with genetic and gene expression variation in HapMap cell lines. Genome Biol 12, R10 (2011).__"

\begin{chapabstract}
Co-methylation is the dependence structure of DNA methylation data. This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation. In this chapter I critically review these definitions and put them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}. I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}. I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.
\end{chapabstract}

@@

I define co-methylation as the dependence structure of DNA methylation data. This broad definition encompasses several similar concepts previously described in the literature. These include: "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha}; "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; the "correlation of two [loci] across many samples" \citep{Akulenko:2013fl}; and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}.

It is important to understand that there exists both within-sample and between-sample co-methylation.

Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample. Many researchers have observed that there is a strong spatial dependence of DNA methylation along the genome, that is, methylation loci near to one another in the genome tend to be similarly methylated \citep[e.g.,][]{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010dr, Landan:2012kp}. These studies also found that the strength of this dependence itself depends on the distance between the loci and on the genomic context.

@@

Between-sample co-methylation is the relationship between methylation levels at a pair of loci across a set of samples[^concordance]. For example, \citet{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples. It is worth noting that measures of within-sample co-methylation can themselves be compared between samples.

[^concordance]: One measure of between-sample correlation of aggregate methylation levels that does not fall under my definition of co-methylation is the following: Consider a pair of samples, $j$ and $j'$, and compute the correlation between the set of $\beta$-values for sample $j$ against sample $j'$, i.e., $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$. This measure is often reported as evidence for the "concordance" of methylation levels between replicates, but is not what I consider a form of co-methylation.

There have been several attempts to analyse co-methylation, most with the aim of better understanding the biological processes that lay down and regulate DNA methylation. An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation. For example, we might identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci. This idea is analogous to the use of tag-SNPs on genotyping microarrays.

@@

\section{Correlations of aggregate methylation levels}
\label{sec:previous_cor_of_agg_methylation}

To date, most analyses of co-methylation have been based on analysing correlatiosn of aggregate methylation levels, such as $\beta$-values, within individual samples \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}. These analyses have an implicit assumption of stationarity, that is, that the correlation between two loci depends only on the distance between them and not the actual position of the loci in the genome.

@@

\citet{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples. Of particular relevance to a discussion of co-methylation, \citet{Eckhardt:2006gh} is often cited as evidence of co-methylation. For example, \citet{Hebestreit:2013ko} cite \citet{Eckhardt:2006gh} as evidence that "methylation levels are strongly spatially correlated" and \citet{Hansen:2011gu} cite \citet{Eckhardt:2006gh} as evidence that"proximal CpGs [have] similar methylation levels".

\citet{Eckhardt:2006gh} performed PCR-amplicon bisulfite-sequencing of $2,524$ amplicons on human chromosomes $6$, $20$ and $22$. PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern whole-genome bisulfite-sequencing assays. The samples are a variety of human tissue types from multiple donors of different ages, both male and female.

As part of their analysis, the authors explore co-methylation, which they define as "the relationship between the degree of methylation over distance". They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $>2,000$ bp" and reference the plot reprinted in Figure \ref{fig:Eckhardt} in support of this claim. In Figure \ref{fig:Eckhardt}, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp. This is not a correlation in the usual sense of the word. Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed[^beck].

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/ng1909-F3a.jpg}
\caption{Correlation of DNA methylation with spatial distance. (a) Correlation between comethylation and spatial distance. Orange dots represent CpG methylation values aggregated and averaged over $25,000$ individual measurements. Gray dots represent CpG methylation values based on resampling of random CpG positions. Blue dots indicate CpG methylation values based on resampling of amplicon positions. At distances $>1,000$ bp, we did not detect any correlation between CpG methylation and spatial distance. Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Eckhardt:2006gh}), copyright (2006)}
\label{fig:Eckhardt}
\end{figure}

[^beck]: An email discussion with the senior author, Professor Stephan Beck (University College London), was unable to clarify the definition or method.

@@

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{i, h}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from distinct samples with different genome-wide average methylation levels. This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h', i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{\ast}$ as follows:

@@

The proof of this result is given in Appendix \ref{app:bernoulli_equal}.

@@

1. $PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, \allowbreak Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.

@@

Quite apart from what exactly is plotted in Figure \ref{fig:Eckhardt}, the analysis is based on data from only $2,524$ amplicons. These limited data cannot give a genome-wide picture of co-methylation. Furthermore, the data come from a variety of tissues and donors, which makes it difficult to know how applicable these results are for a particular tissue. Caution is warranted in extrapolating these results to general statements about the correlation of aggregate methylation levels.

@@

\citet{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing. They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis thaliana_. They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis]. The autocorrelation computes, for a set of distances ($0-5,000$ bp), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance. To emphasise, I believe this analysis uses all pairs of methylation loci, that is, regardless of how many intervening methylation loci there are for each pair.

@@

The autocorrelation plots showed two periodicities, one of $\approx 167$ bp and the other of $10$ bp (CHH only). Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier analysis of the autocorrelation signal.

[^CHH_only]: The $10$ nucleotide period was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}

The $167$ bp period is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al. hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.

The $10$ bp period is equal to the length of one helical DNA turn. From this result Cokus et al. hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites $8$ to $10$ bp apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_). The $10$ bp CHH-only period in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})

[^comethylation_def2]: This is also referred to as "co-methylation" by \citet{Jurkowska:2011dt}, here meaning the methylation of multiple loci in a single event. \citet{Jurkowska:2011dt} review a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8$ to $10$ bp; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.

\citet{Lister:2009hy} report an $8$ to $10$ bp period in the occurrence "methylcytosines". Recall that \citeauthor{Lister:2009hy} define a "methylcytosine" as a cytosine that has a non-zero $\beta$-value, as determined by a binomial test (see Section \ref{sec:beta-value_interpretation} for details and a criticism of this concept). With its reference to "methylcytosines", this may first appear to be a within-fragment analysis of co-methylation, but it is in fact based on $\beta$-values. However, unlike other methods described in this section, the analysis is not based on a correlation of these $\beta$-values, but rather on patterns in the spacing of "methylcytosines" along the genome. This leads to problems.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/Lister_2009_Sup_Fig_9.pdf}
\caption{Spacing of Adjacent Methylcytosines in Different Contexts. Prevalence of mCHG/mCHH sites (y-axis) as a function of the number of bases between adjacent mCHG/mCHH sites (x-axis) based on all non redundant pair-wise distances up to $50$ nt in exons, introns and random sequences. The blue line represents smoothing by cubic splines. Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Lister:2009hy}), copyright (2009)}
\label{fig:Lister_2009_Sup_Fig_9}
\end{figure}

An $8$ to $10$ bp period in the fitted cubic spline is observed in most, but not all, of these nine plots. This periodicity is very weak in several contexts and is absent for exonic CpGs. The weakness of the signal is partially attributable to the small number of "methylcytosines" in the CHG and CHH contexts in human samples. More importantly, the reported $8$ to $10$ bp periods are weaker than a $3$ bp period for "methylcytosines" at exonic CpGs. Unlike the $8$ to $10$ nucleotide period, the $3$ nucleotide period can clearly be observed in the raw counts, although the cubic spline smoother seemingly ignores it. \citeauthor{Lister:2009hy} do not comment on the $3$ bp period.

Based on these nine plots, \citet{Lister:2009hy} draw parallels to the $10$ bp period in the autocorrelation of CHH $\beta$-values observed by \citet{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites". However, I suspect that the results in these nine plots are in fact driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines. That is not to say that there aren't periodicities in the methylation of cytosines, but that is not what is being measured in this analysis. To explain, these plots use the raw counts of "methylcytosines", which are not adjusted for the frequency of pairs of cytosines with a given $IPD$. If pairs of cytosines are observed more often at certain $IPD$s then even if methylation was totally independent of $IPD$ we would expect more pairs of "methylcytosines" at these same $IPD$s. Therefore, at a minimum, it is necessary to normalise the number of "methylcytosines" by the number of cytosines.

In fact, there are well known periodicities in the distribution of nucleotides in genomes. For example, in eukaryotic genomes there is a $10$ bp period of G+C dinucleotides that acts out of phase with a $10$ bp period of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8$ to $10$ nucleotide period and the latter the $3$ bp period for exonic CpGs.

\subsection{\citet{Chodavarapu:2010iqa}}\label{sec:chodavarapu}

\citet{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (_HSF1_). The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq. The nucleosome start sites anchor the methylation observations to a common grid and this plot is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes. A $10$ bp periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}. Again, the observed periodicity is confirmed by analysing a Fourier transform of this correlation signal.

In addition to the genome-wide analyses, the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of the _Arabidopsis_ sample and promoters, genes, repeats and CpG islands of the _HSF1_ sample, which confirmed the $10$ bp periodicity. In fact, these results revealed that the $10$ bp period is common to all methylation types, including symmetric CpG methylation, at least for nucleosome-bound DNA. This led \citeauthor{Chodavarapu:2010iqa} to discard their earlier hypothesis[^same_authors] as to the source of the $10$ bp period \citep{Cokus:2008fc} in favour of one that "nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__ [emphasis added]" and not just for the DRM2 methyltransferase.

[^same_authors]: Many of the same authors contributed to \citet{Cokus:2008fc} and \citet{Chodavarapu:2010iqa}.

@@

\citet{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation". \citeauthor{Li:2010fb} conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", and that their results are consistent with those in \citet{Eckhardt:2006gh}. In addition to this genome-wide analysis, \citet{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated regions, coding DNA sequences and various repetitive elements. Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.

@@

Figure \ref{fig:Li6} plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are estimates of the correlation of aggregate methylation levels over the range $0$ to $1000$ bp. Based on the "$n = 18,936,995$" quoted in the header of panel (a) of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "distance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs. If all pairs were used, $n$ would be much larger since there are more than twenty million CpGs on one strand of the haploid human genome. To emphasise, the correlation computed by \citet{Li:2010fb} is not the same as the autocorrelation computed in \citet{Cokus:2008fc} since \citet{Li:2010fb} only use pairs of adjacent CpGs whereas \citet{Cokus:2008fc} use all pairs of CpGs, regardless of the number of intervening CpGs.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/{journal.pbio.1000533.s006}.pdf}
\caption{Co-methylation patterns for different genomic features. A reproduction of Supplementary Figure 6 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
\label{fig:Li6}
\end{figure}

The plot of genome-wide correlations (panel (a) in Figure \ref{fig:Li6}) contains a period of approximately $170$ bp, very similar to the $167$ bp period identified in _Arabidopsis_ by \citet{Cokus:2008fc}. This is further evidence that nucleosome spacing effects the strength of co-methylation. However, \citet{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ bp period reported by \citet{Cokus:2008fc} and \citet{Chodavarapu:2010iqa}.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/Figure_S7.pdf}
\caption{Transformation of methylation correlation at nearby CpG cytosines. A reproduction of Supplementary Figure 7 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
\label{fig:Li7}
\end{figure}

@@

Figure \ref{fig:Lyko} reproduces the results for the co-methylation analysis. Note that the maximum value of the "CpG autocorrelation" is less than $0.015$ (panel (A) of Figure \ref{fig:Lyko}). I strongly suspect that the autocorrelation has been incorrectly computed. Specifically, what I believe has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been artificially assigned a $\beta$-value of zero. It does not make sense to include non-CpG positions, particularly non-cytosines that can never be methylated, in an analysis of CpG co-methylation. I do not have the data used in \citet{Lyko:2010dr}, but the following example supports my conclusion. This example uses the chromosome 22 data from the _FF_ sample in the _Lister_ dataset (see Section \ref{sec:Lister_dataset} for full details of this sample).

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/Lyko_Figure_S7.pdf}
\caption{Periodicity of methylation patterns. (A) Autocorrelation of CpG methylation status over 1 kb. (B) Autocorrelation over 100 bp. Figures A and B show that the correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away. CpGs within a neighborhood of 2 bp to 100 bp are thus more likely to share the same methylation status than more distant CpGs. (C) Fourrier transform of autocorrelation showing a clear periodicity peak at 33 cycles per 100 bp (every 3 bp). (D) Distribution of codon position of mCs, and distribution of methylation level depending on the position. These two panels indicate that the distance between methylated CpGs is often a multiple of three and that the methylated cytosine corresponds most frequently to the first nucleotide of an arginine codon. A reproduction of Supplementary Figure 7 from \citet{Lyko:2010dr} under the Creative Common Attribution (CCBY) licence.}
\label{fig:Lyko}
\end{figure}

Let $\bm{\beta^{\ast}}$ be the augmented vector of $\beta$-values, where:
